Structure Learning for $\VAR(p)$ Models

### 2019-06-10 paper/paper.tex
Updated paper with hypothesis test section and algorithm to match code.

\begin{definition}[Granger Causality]
  For the WSS series $x(t)$ we say that component $x_j$
  \textit{Granger-Causes} (GC) component $x_i$ (with respect to $x$)
  and write $x_j \gc x_i$ if given Hilbert spaces
  $\H_{t - 1}$, $\H^{-j}_{t - 1}$

@@

@@

\subsection{Layered Strongly Causal Graphs}
It is clearn that Algorithm \ref{alg:pwgr} is a purely theoretical construct, and cannot be applied directly to real data.  Instead, we convert the algorithm into an heuristic and observe it's performance experimentally.  We first consider the heuristic version of Algorithm \ref{alg:pwgr} in Algorithm \ref{alg:pwgr_heuristic} and show that is a well defined procedure:

@@

  \BlankLine
  \caption{Graph Recovery Heuristic $\mathsf{pw\_scg}(F, \delta, b, R)$}
  \label{alg:pwgr_heuristic}
  % \TitleOfAlgo{Pairwise Graph Recovery}
  \Input{Estimates of pairwise Granger-causaltiy statistics $F_{ij} = \ln \frac{\xi_{ij}}{\xi_i}$\\
    Sparsity parameters ($\delta, b, R$)}
  \Output{A strongly causal graph $\widehat{\gcg}$}
  \Initialize{$S = [n]$  \texttt{\# unprocessed nodes}\\
    $E = \emptyset$  \texttt{\# edges of }$\widehat{\gcg}$\\
    $k = 1$ \texttt{\# a counter used only for notation}}

@@

  $W_\delta \leftarrow \{(i, j)\ |\ F_{ji} \ge \delta, F_{ji} > F_{ij}\}$  \texttt{\# candidate edges}\\
  $\mathcal{I}_0 \leftarrow \big(\sum_{j\in S: (j, i) \in W_\delta} F_{ij}, \mathsf{\ for\ }i \in S \big)$ \texttt{\# total incident strength for each node}\\
  $P_0 \leftarrow \mathsf{arg\_select\_min\_N}(\mathcal{I}_0, N=R)$  \texttt{\# select} $R$ \texttt{root nodes}\\
  \BlankLine

  \While{$S \ne \emptyset$}{
    $S \leftarrow S \setminus P_{k - 1}$ \texttt{\# remove processed nodes}\\
    $\mathcal{I}_k \leftarrow \big(\sum_{j \in S: (j, i) \in W_\delta} F_{ij}, \mathsf{\ for\ }i \in S \big)$ \texttt{\# intra-}$S$ \texttt{incident strength}\\
    $P_k \leftarrow \mathsf{arg\_select\_min\_N}(\mathcal{I}_k, N=b)$  \texttt{\# select} $b$ \texttt{branching nodes}\\
    \;
    \texttt{\# add strongest} $P_{k - r} \rightarrow P_k$ \texttt{edges and maintain strong causality}\\
    \For{$r = 1, \ldots, k$} 
    {
      \For{$(i, j) \in \mathsf{sort}\Big(\{(i, j) \in P_{k - r} \times P_k\ |\ (i, j) \in W_\delta\} \mathsf{\ by\ descending\ } F_{ji}\Big)$}{
        \If{no $\gcgpath{u}{j}$ path in $E$, for any $u \in \anc{i}$}{
          $E \leftarrow E \cup \{(i, j)\}$
        }
      }
    }
    $k \leftarrow k + 1$
  }
  \Return{$([n], E)$}
\end{algorithm}

>>>>

\begin{definition}[Granger Causality]
  \label{def:granger_causality}
  For the WSS series $x(t)$ we say that component $x_j$
  \textit{Granger-Causes} (GC) component $x_i$ (with respect to $x$)
  and write $x_j \gc x_i$ if given Hilbert spaces
  $\H_{t - 1}$, $\H^{-j}_{t - 1}$

@@

\section{Finite Sample Graph Recovery}
\label{sec:structure_learning}
In this section we provide a review of our methods for implementing
Algorithm 1 given a \textit{finite} sample of $T$ data points.  We
apply the simplest reasonable methods in order to maintain a focus on
our main contributions (i.e. Algorithm \ref{alg:pwgr}), more
sophisticated schemes can only serve to improve the results.  Textbook
reviews of the following concepts are provided e.g. by
\cite{all_of_statistics}, \cite{murphy_mlp}, and elsewhere.

We proceed first by defining pairwise Granger-causality hypothesis
tests, second a model order selection criteria, and finally the method
for choosing a $P-$value cutoff.

\subsection{Hypothesis Testing}
In performing pairwise checks for Granger-causality $x_j \pwgc x_i$ we
follow the simple scheme of estimating the two linear models via
ordinary least squares:

\begin{align}
  H_0:&\ \widehat{x}_i^{(p_i)}(t) = \sum_{\tau = 1}^{p_i} b_{ii}(\tau)x_i(t - \tau),\\
  H_1:&\ \widehat{x}_i^{(p_i, p_j)}(t) = \sum_{\tau = 1}^{p_i} b_{ii}(\tau)x_i(t - \tau) + \sum_{\tau = 1}^{p_j}b_{ij}(\tau)x_j(t - \tau).
\end{align}

Under the null hypothesis $H_0$ (i.e. that $x_j \npwgc x_i$) we can
form the asymptotically $\chi^2(p_j)$ distributed test statistic
(\hl{TODO: Citation that this is chi2})

\begin{equation}
  \label{eqn:gc_statistics}
  F_{ij}(p_i, p_j) = \frac{T}{p_j}\Big(\frac{\xi_i(p_i)}{\xi_{ij}(p_i, p_j))} - 1\Big),
\end{equation}

where $\xi_i(p_i)$ is the sample mean square of the residuals\footnote{This
  quantity is often denoted $\widehat{\sigma}$, but we maintain
  notation from Definition \ref{def:granger_causality}.}
$x_i(t) - \widehat{x}^{(p_i)}_i(t)$

\begin{equation*}
  \xi_i(p_i) = \frac{1}{T - p_i}\sum_{t = p_i + 1}^T (x_i(t) - \widehat{x}_i^{(p_i)}(t))^2
\end{equation*}

Given additional assumptions on $x(t)$ (\hl{TODO: What assumptions? --
  this must also depend on choice of $p_i, p_j$}) we have the
convergence (\hl{a.s. or in P?  -- I think in P})

\begin{equation}
  \underset{T \rightarrow \infty}{\text{lim }}F_{ij} =
  \left\{
    \begin{array}{ll}
      0;\ x_j \npwgc x_i\\
      \infty;\ x_j \pwgc x_i
    \end{array}
  \right.
\end{equation}

Therefore, we can make the statement that Algorithm
\ref{alg:pwgr_heuristic} asymptotically implements Algorithm
\ref{alg:pwgr}.  \hl{TODO: Modify Algorithm 2 to make this true.}
\hl{TODO: Enclose this into a theorem}

\subsection{Model Order Selection}
\label{sec:model_order_selection}
There are a variety of methods to choose the filter orders $p_i$ and
$p_j$ \hl{TODO: cite some more}, we will focus in particular on the
Bayesian Information Criteria (BIC).  The BIC is substantially more
conservative than the Akaiake Information Criteria, and since we are
searching for \textit{sparse graphs}, we therefore prefer the BIC. For
the bivariate case we focus for simplicity on $p_i = p_j$ and write
simply\footnote{We use the same $p$ for the bivariate case, but this
  need not be the same as the corresponding univariate estimate}
$x_{ij}^{(p)}$, in which case the BIC, which we seek to
\textit{maximize} over $p$ is given by:

\begin{equation}
  \label{eqn:bic}
  \begin{aligned}
    BIC_{\text{univariate}}(p) &= -(T - p - 1) \ln \xi_i(p) - p\ln T,\\
    BIC_{\text{bivariate}}(p) &= -(T - p - 1) \ln \xi_{ij}(p) - 2p\ln T.\\
  \end{aligned}
\end{equation}

We carry this out via the forward search described in Algorithm
\ref{alg:maximize_bic}.  Since $BIC(p)$ need not be a convex function,
this scheme is not guaranteed to find the true maximum of the BIC, but
this approximation suffices for our purposes \hl{TODO: is this
  sufficient?  A full search is not so much harder.}

\begin{algorithm}
  \caption{BIC Model Selection Algorithm}
  \label{alg:maximize_bic}

  \SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
  \SetKwInOut{Initialize}{initialize}
  \DontPrintSemicolon

  \BlankLine
  \Input{$T$ samples of WSS processes $x_i(t), x_j(t)$.  A maximum possible lag $p_{\text{max}}$.}
  \Output{Model orders $p_i$ and $p$ for models in Hypotheses $H_0$ and $H_1$ respectively.}
  \Initialize{
    $BIC_{\text{univariate}} = -\infty$\\
    $BIC_{\text{bivariate}} = -\infty$\\
  }

  \# \texttt{Find $p_i$}\\
  \For{$p \in \{1, 2, \ldots, p_{\text{max}}\}$}{
    $\xi_i(p) = $\ \texttt{fit\_univariate}$(x_i(t), p)$ \texttt{ \# Fit model for $H_0$}\\
    $bic_p = -(T - p - 1) \ln \xi_i(p) - p\ln(T - p - 1)$\\
    \If{$bic_p > BIC_{\text{univariate}}$}{
      $BIC_{\text{univariate}} = bic_p$\\
      \texttt{continue}\\
    } \Else{
      $p_i = p - 1$\\
      \texttt{break}\\
    }
  }

  \BlankLine
  \# \texttt{Find $p$}\\
  \For{$p \in \{1, 2, \ldots, p_{\text{max}}\}$}{
    $\xi_{ij}(p) = $\ \texttt{fit\_biivariate}$(x_i(t), x_j(t), p)$ \texttt{ \# Fit model for $H_1$}\\
    $bic_p = -(T - p - 1) \ln \xi_{ij}(p) - 2p\ln(T - p - 1)$\\
    \If{$bic_p > BIC_{\text{bivariate}}$}{
      $BIC_{\text{biivariate}} = bic_p$\\
      \texttt{continue}\\
    }\Else{
      $p = p - 1$\\
      \texttt{break}\\
    }
  }
    \Return{$p_i, p$}
\end{algorithm}

\subsection{Edge Probabilities and Error Rate Controls}
Denote $F_{ij}$ the Granger-causality statistic of equation
\ref{eqn:gc_statistics} with model orders chosen by the methods of
Section \ref{sec:model_order_selection}.  Note that this statistic is
asymptotically $\chi^2(p)$ distributed, and denote by $G$ the
$\chi^2(p)$ cumulative distribution function.  We will define the matrix

\begin{equation}
  \label{eqn:edge_inclusion_probability}
  P_{ij} = G(F_{ij}),
\end{equation}

to be the matrix of pairwise edge inclusion probabilities.  This is
motivated by the hypothesis test where the hypothesis $H_0$ will be
rejected (and thence we will conclude that $x_j \pwgc x_i$) if
$P_{ij} > 1 - \delta$.

The value $\delta$ can be chosen by a variety of methods, in our case
we apply the Benjamini Hochberg criteria \cite{benjamini_hochberg}
\cite{all_of_statistics} to control the false discovery rate of
pairwise edges to a level $\alpha$ (where we generally take
$\alpha = 0.05$).

@@

\subsection{Layered Strongly Causal Graphs}
It is clearn that Algorithm \ref{alg:pwgr} is a purely theoretical construct, and cannot be applied directly to real data.  Instead, we convert the algorithm into an heuristic and observe it's performance experimentally.  We first consider the heuristic version of Algorithm \ref{alg:pwgr} in Algorithm \ref{alg:pwgr_heuristic} and show that is a well defined procedure (note that $\ceil{x}$ denotes the integer ceiling function):

@@

  \BlankLine
  \caption{Graph Recovery Heuristic $\mathsf{pw\_scg}(F, P, \alpha)$}
  \label{alg:pwgr_heuristic}
  \Input{Estimates of pairwise Granger-causaltiy statistics $F_{ij}$ (eqn \ref{eqn:gc_statistics}), Matrix of edge probabilities $P_{ij}$ (eqn \ref{eqn:edge_inclusion_probability}), threshold $\delta$.}
  \Output{A strongly causal graph $\widehat{\gcg}$}
  \Initialize{$S = [n]$  \texttt{\# unprocessed nodes}\\
    $E = \emptyset$  \texttt{\# edges of }$\widehat{\gcg}$\\
    $k = 1$ \texttt{\# a counter used only for notation}}

@@

  $W_\delta \leftarrow \{(i, j)\ |\ P_{ji} > 1 - \delta, F_{ji} > F_{ij}\}$  \texttt{\# candidate edges}\\
  $\mathcal{I}_0 \leftarrow \big(\sum_{j\in S: (j, i) \in W_\delta} P_{ij}, \mathsf{\ for\ }i \in S \big)$ \texttt{\# total node incident probability}\\
  $P_0 \leftarrow \{i \in S\ |\ \mathcal{I}_0(i) < \ceil{\text{min}(\mathcal{I}_0)}\}$ \texttt{\# Nodes with fewest incident edges}\\
  \If{$P_0 = \emptyset$}{
    $P_0 \leftarrow \{i \in S\ |\ \mathcal{I}_0(i) \le \ceil{\text{min}(\mathcal{I}_0)}\}$ \texttt{\# Ensure non-empty}
  }
  \BlankLine

  \While{$S \ne \emptyset$}{
    $S \leftarrow S \setminus P_{k - 1}$ \texttt{\# remove processed nodes}\\
    % $\mathcal{I}_k \leftarrow \big(\sum_{j \in S: (j, i) \in W_\delta} F_{ij}, \mathsf{\ for\ }i \in S \big)$ \texttt{\# intra-}$S$ \texttt{incident strength}\\
    $\mathcal{I}_k \leftarrow \big(\sum_{j\in S: (j, i) \in W_\delta} P_{ij}, \mathsf{\ for\ }i \in S \big)$\\
    $P_k \leftarrow \{i \in S\ |\ \mathcal{I}_k(i) < \ceil{\text{min}(\mathcal{I}_k)}\}$\\
    \If{$P_k = \emptyset$}{
      $P_k \leftarrow \{i \in S\ |\ \mathcal{I}_k(i) \le \ceil{\text{min}(\mathcal{I}_k)}\}$
    }
    \;
    \texttt{\# add strongest edges, maintaining strong causality}\\
    $U_k \leftarrow \bigcup_{r = 1}^k P_{k - r}$ \texttt{\# Include all forward edges}\\
    \For{$(i, j) \in \mathsf{sort}\Big(\{(i, j) \in U_k \times P_k\ |\ (i, j) \in W_\delta\} \mathsf{\ by\ descending\ } F_{ji}\Big)$} {
      \If{$\mathsf{is\_strongly\_causal}(E \cup \{(i, j)\})$} {
        $E \leftarrow E \cup \{(i, j)\}$
      }
    }
    $k \leftarrow k + 1$\\
  }
  \Return{$([n], E)$}
\end{algorithm}
